Code Test Part 2: Baby Names!
========================================================

The analysis could be recreated by running the file `run.R` in R by `source("run.R")`.
The data are from [http://www.ssa.gov/oact/babynames/state/namesbystate.zip](http://www.ssa.gov/oact/babynames/state/namesbystate.zip).


Load Data
----------
The data set is loaded with the following code.
```{r}
# Get the data from the web
fileUrl <- "http://www.ssa.gov/oact/babynames/state/namesbystate.zip"
download.file(fileUrl, destfile = "namesbystate.zip")
unzip("namesbystate.zip")
# batch read files
files <- list.files(pattern = ".TXT$")
babyname <- lapply(files, read.table, sep = ",")
babyname <- do.call(rbind,babyname)
# Set column names
names(babyname) <- c("state", "gender", "year", "name", "count")
```

For later analysis, I identify states for each region in the United States.
```{r}
NorthWest <- c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA")
MidWest <- c("IL", "IN", "MI", "OH", "WI", "IA", "KS", "MN", "MO", "NE", "ND", "SD")
South <- c("DE", "FL", "GA", "MD", "NC", "SC", "VA", "DC", "WV", "AL", "KY", "MS", "TN", "AR", "LA", "OK", "TX")
West <- c("AZ", "CO", "ID", "MT", "NV", "NM", "UT", "WY", "AK", "CA", "HI", "OR", "WA")
```

And load all packages needed.
```{r}
library(plyr)
library(dplyr)
library(tidyr)
library(data.table)
library(reshape2)
library(ggplot2)
```

Let's check some summary statistics.
```{r}
str(babyname)
summary(babyname)
```


Descriptive Analysis
--------------------

### Question1
Please describe the format of the data files. Can you identify any limitations or distortions of the data?
The files are in CSV format and each record is separated by a comma. If the record contains comma, CSV format 
cannot distinguish the record and the delimiter. As described in [Social Security website](http://www.ssa.gov/oact
/babynames/limits.html), this particular data set contains only names with at least 5 occurrences in each state.
Therefore, if a state has many unique names with less than 5 occurrences, then the popularity of a name for a year 
in that state (calculated by the occurrences of that name in a year divided by the total number of occurrences 
of any name in a year) will be biased. In addition, the data set contains only names with length between 2 and 15.
And it does not contain names with special characters (e.g. Amélie). It is unclear how special characters are 
treated in this data set. 

### Question 2
What is the most popular name of all time? (Of either gender.)
This can be done by apply the `sum` function to count of names by each name.
```{r}
head(sort(tapply(babyname$count, babyname$name, sum), decreasing = TRUE))
```
The most popular name of all time is James with 4,942,431 occurrences, followed by John with 4,834,422 occurrences.

### Question 3
What is the most gender ambiguous name in 2013? 1945?
I wrote a function to answer this quesiton. This function compute the proportion of female/male for a given 
name. The proportion of female for a given name is called propF in the function. To compare gender ratio, 
I compute the absolute difference between propF and 0.5, where 0.5 is the case when equal number of male/female
have the given name. This absolute difference is called deviance in the function, and the smaller the deviance,
the more gender-neutral the name is (more gender ambiguous).
```{r}
genderNeutral <- function(data, year) {
  #subset data
  data <- data[(data$year == year),]
  #combine counts from different states and drop year
  data <- aggregate(data$count, by = list(data$year, data$name, data$gender), FUN = sum)
  data <- data[,-1]
  names(data) <- c("name", "gender", "count")
  # spread the data for two gender level
  byName <- spread(data, gender, count)
  # replace NA with 0
  byName[is.na(byName)] <- 0
  # create new variable that records proportion of female for each name
  byName <- mutate(byName, propF = F/(F+M), deviance = abs(F/(F+M) - 0.5))
  print(head(arrange(byName,deviance, desc(F)), n = 10))
}
```    

And then I applied the function to year 1945 and 2013.
The most gender ambiguous names for 2013 are: 
```{r}
genderNeutral(babyname, 2013)
```

The most gender ambiguous names for 1945 are: 
```{r}
genderNeutral(babyname, 1945)
```

This function can be applied to any given year.


### Question 4
Of the names represented in the data, find the name that has had the largest percentage increase in popularity since 1980. Largest decrease?

I first created a new data set **allYear** that contains year, name, count, total number of names for the year,
and proportion of the name relative to all names in that year. The last variable `prop` reflects the popularity
of a name in that year. The higher the proportion, the more popular the name in that year.
```{r}
# combine counts from different states and different gender
allYear <- aggregate(babyname$count, by = list(babyname$year, babyname$name), FUN = sum)
names(allYear) <- c("year", "name", "count")

# compute the total number of names for each year
countByYear <- ddply(allYear, .(year), summarise, total = sum(count))
# join allYear with countByYear by "year"
allYear <- merge(allYear, countByYear, by.x = "year", by.y = "year")
# create new variable: proportion of a given name relative to all names in that year
allYear <- mutate(allYear, prop = count / total)
```

And let's take a look at *allYear*.
```{r}
head(allYear)
```

Now let's compare the popularity of names in 1980 and 2013.
I select year 1980 and 2013 from *allYear*, and inner join these two data sets by name.
This will only keep names that are present in both years.
Then I create a new variable `prop.diff` that computes the difference between popularity.

```{r}
# compute differences of prop between 1980 and 2013 for each name.
year1980 <- allYear[(allYear$year == 1980),]
year2013 <- allYear[(allYear$year == 2013),]
# inner join two data sets using name
combineTable <- merge(year1980, year2013, by.x = "name", by.y = "name")
combineTable <- mutate(combineTable, prop.diff = prop.y - prop.x)
```

The top 10 names that has had the largest percentage increase in popularity since 1980 are:
```{r}
# Top 10 names increase in popularity
head(arrange(combineTable, desc(prop.diff)), n = 10)
```

And the top 10 names that has had the largest percentage decrease in popularity since 1980 are:
```{r}
# Top 10 names decrease in popularity
head(arrange(combineTable, desc(-prop.diff)), n = 10)
```

### Question 5
Can you identify names that may have had an even larger increase or decrease in popularity?

Utilizing the *allYear* data set, we can repeat the same procedure in 4. 
Now let's compare year 1910 and 2013.
```{r}
year1910 <- allYear[(allYear$year == 1910),]
# inner join two data sets using name
combineTable2 <- merge(year1910, year2013, by.x = "name", by.y = "name")
combineTable2 <- mutate(combineTable2, prop.diff = prop.y - prop.x)
```

The top 10 names that has had the largest percentage increase in popularity since 1910 are:
```{r}
head(arrange(combineTable2, desc(prop.diff)), n = 10)
topNameIncrease <- arrange(combineTable2, desc(prop.diff))$name[1:10]
```
The largest percentage increase in popularity since 1910 is smaller than that since 1980.

Let's check the top 10 names that has had the largest percentage decrease in popularity since 1910.
```{r}
head(arrange(combineTable2, desc(-prop.diff)), n = 10)
topNameDecrease <- arrange(combineTable2, desc(-prop.diff))$name[1:10]
```
This shows that **Mary**, **Helen** and **John** have had even larger decrease in popularity since 1910 than
**Jennifer** since 1980.

Now let's try to plot the trend of these top 10 name increase/decrease in popularity.
In order to do that, I wrote two functions to plot trend in popularity for a given name.
And the multiplot function is from [Cookbook for R](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/).
```{r}
## This function plot the trend for a given name
plotNameTrend <- function(data, name) {
  nameData <- data[(data$name == name),]
  g <- ggplot(nameData, aes(year, prop)) + geom_line() + labs(x = "Year", y = "Popularity", title = nameData$name)
  return(g)
}

## This function create a multiplot for a list of name using the plotNameTrend() function.
plotNameList <- function(data, nameList) {
  plots <- list()
  for (i in seq(along = nameList)) {
    g <- plotNameTrend(data, nameList[i])  
    plots[[i]] <- g
  }
  multiplot(plotlist = plots, cols = 2)
}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

Then we can check the trend of those top 10 names that have had the largest percentage increase/decrease in
popularity.
```{r}
# drop count and total
propAllYear <- select(allYear, -count, -total)

plotNameList(propAllYear, topNameIncrease)
plotNameList(propAllYear, topNameDecrease)
```

Onward to Insight!
------------------
There are many interesting questions can be asked for this data set. Here I tried to address some of them.

1. Do people prefer longer names nowadays?

To answer this question, I utilize the **propAllYear** data to create a new data set **nameLength** that record
the popularity for each name length for each year

```{r}
propAllYear$name <- as.character(propAllYear$name)
# create new variable: length of each name
nameLength <- mutate(propAllYear, nchar = nchar(name))
# delete name column
nameLength <- select(nameLength, -name)
# combine prop for the same length of name for each year
nameLength <- ddply(nameLength, .(year, nchar), summarise, total = sum(prop))
head(nameLength)
```

Then I plot the trend by each number of characters in name. Since the popularity of each name length differs 
a lot, I also create individual plots.
```{r}
# plot the name length trend 
plots <- list()
g <- ggplot(nameLength, aes(year, total, group = nchar, color = nchar))
plots[[1]] <- g + geom_line() + scale_color_gradient2(midpoint = 9, low = "blue", mid = "white", high = "red")

# plot the trend for each name length (easier to check the trend)
for (i in 2:15) {
  nameLength_sub <- nameLength[(nameLength$nchar == i),]
  myTitle <- paste("Name length = ", i)
  g <- ggplot(nameLength_sub, aes(year, total)) + geom_line() + labs(x = "Year", y = "Popularity", title = myTitle)
  plots[[i]] <- g
}
multiplot(plotlist = plots, cols = 3)
```

It shows that for very short name (length = 2), the popularity increases from 1925, and decreases around 1955.
Shorter names seem to decrease in general since 1910, and gradually increase since around 2000.
Names with length between 4 and 8 are more common than others. 
And since 1960, longer names seem to increase in popularity, especially length = 9, 11, 12, 13.

2. Are there any differences in name popularity between regions?

I wrote a function to gather the top names for a region in a given year.
```{r}
## This function find the top names for a year in a given region.
topName <- function(data, region, year, top) {
  temp <- data[(data$state %in% region & data$year == year),]
  temp <- ddply(temp, .(name), summarise, total = sum(count))
  temp <- arrange(temp, desc(total))[1:top,]
  return(temp)
}
```

Then let's find the top 5 names in 2013 for each region.
```{r}
top5_NW2013 <- topName(babyname, NorthWest, 2013, 5)
top5_MW2013 <- topName(babyname, MidWest, 2013, 5)
top5_S2013 <- topName(babyname, South, 2013, 5)
top5_W2013 <- topName(babyname, West, 2013, 5)

region <- c("NorthWest", "MidWest", "South", "West")
plots <- list()
tops <- list(top5_NW2013, top5_MW2013, top5_S2013, top5_W2013)
for (i in seq(along = region)) {
  myTitle <- region[i]
  g <- ggplot(tops[[i]], aes(x = factor(name), y = total, fill = name)) + geom_bar(stat = "identity") + labs(x = "Name", y = "Count", title = myTitle)
  plots[[i]] <- g
}
multiplot(plotlist = plots, cols = 2)
```

**Emma**, **Sophia** are top 5 in all regions. **Isabella** is in top 5 in North West and in West region. 
Is **Isabella** going to be the next **Sophia**?
And there are other names (**Liam**, **Olivia**, **Manson**, **Noah**) that are top 5 in two regions. 
**William** is in top 5 in South region but not in other regions. 
**Jacaob** is in top 5 in West region but not in other regions.

3. Does New York or California lead the trend of names?

To answer this question, I utilize the previous `topName()` function
```{r}
top5_NY <- topName(babyname, "NY", 2013, 5)
top5_CA <- topName(babyname, "CA", 2013, 5)
top5_NY
top5_CA
```
The top 2 popular name in New York and California are both Sophia and Jacob.
Let's check the trend of these names in NY and CA. 
For comparison, I also find the trend in North Carolina (NC).

First, I write two more functions to compute the popularity of a name in a state and plot trends of name.
```{r}
## This function compute the popularity of a name in a state 
propNameInState <- function(data, state) {
  temp <- select(data[(data$state == state),], -state)

  # combine counts from different gender
  temp <- aggregate(temp$count, by = list(temp$year, temp$name), FUN = sum)
  names(temp) <- c("year", "name", "count")
  
  # compute the total number of names for each year in NY
  temp_countByYear <- ddply(temp, .(year), summarise, total = sum(count))
  # join allYear with countByYear by "year"
  temp <- merge(temp, temp_countByYear, by.x = "year", by.y = "year")
  # create new variable: proportion of a given name relative to all names in that year
  temp <- mutate(temp , prop = count / total)
  return(temp)
}

## This function plot the name trend for three given states.
plotNameTrendInState <- function(data1, data2, data3, state1, state2, state3, name) {
  temp1 <- data1[(data1$name == name),]
  temp1$state <- state1
  temp2 <- data2[(data2$name == name),]
  temp2$state <- state2
  temp3 <- data3[(data3$name == name),]
  temp3$state <- state3
  temp <- rbind(temp1, temp2, temp3)
  
  myTitle <- paste("Trend in", name)
  g <- ggplot(temp, aes(year, prop, group = state, color = state)) + geom_line() + labs(x = "Year", y = "Popularity in state", title = myTitle)
  return(g)
}
```

Then I create the plot of trend in Sophia and Jacob for these three states.
```{r}
NY_data <- propNameInState(babyname, "NY")
CA_data <- propNameInState(babyname, "CA")
NC_data <- propNameInState(babyname, "NC")

plotNameTrendInState(NY_data, CA_data, NC_data, "NY", "CA", "NC", "Sophia")
plotNameTrendInState(NY_data, CA_data, NC_data, "NY", "CA", "NC", "Jacob")
```

It seems that the trend of popularity in **Sophia** is quite similar between these three states.
However, it seems that around 1970, **Jacob** became very popular in California than other two states.


There are other questions that I would like to ask. Here are some of them.
- Can you predict someone's age when all you know is the name?
- Inspired by Freakonomics which lists the most "white" and most "black" names, I would like to know
   the most "Asian" names. Are there English names that Asian people prefer?
- Do parents name their children after celebrity? NCAA MVP?
- Are there any boys' name that becomes girls' name?
- How is the prediction of Freaknomics for most popular names of 2015?
